---
title: "Overview of data collected in the field season of 2024 for coastal monitoring"
author: "Maris Pärn"
format: html
editor: visual
---

```{r}
# read necessary data in

overview_summary <- read.csv("overview_2024.csv")

```

## Summary of 2024

### Year in numbers

Number of detectors: 50\

Number of detectors that survived the season: 48\

Reasons of detector failure: Water damage by humans, seawater damage by other forces\

Date of first deployment:\

Date of last deployment:\

Date of first retrieval: `r min(overview_summary$date_retrieved)`\

Date of last retrieval: `r max(overview_summary$date_retrieved)`\

Days detectors were on the field: 4908\

Days with missing data: 320 or 319\

\% of missing data from all data: ca 6,5\

Reasons of missing data:\

1\) No bats, no noise detected: 19\

2\) SD card full: 238\

Batteries empty: 23\

Detector dead: 33\

Average time detector deployed: mean(overview_data\$`days deployed`)\

Time difference of 98.17697 days

```{r}



```

Raw data, auto-id

## Operational

time, comments from survey123, what SDcard problems were had, batterylife etc

```{r}
1 + 1


```

## Noise

How much, when, what happened during daytime?

```{r}
#| echo: false
2 * 2
```

## Bats

Who, when, daytime recordings? - this should not be done with autoid data.
